Gradual Soundness: Lessons from Static Python

### 2021-10-25 ecoop-2022/paper.tex

\begin{abstract}
  Soundness, migratory typing.
  Novel run-time check strategy, combines concrete and transient for ergonomics.
\end{abstract}

@@

We have written this paper with two audiences in mind.
First, we want to encourage system-builders to reproduce the compelling
\SP{} language design in another context and measure performance.
Second, we want to lend focus to theoretical research on gradual typing.
The restrictions that \SP{} adopts (relative to the ideal promise of gradual typing)
may be worthwhile assumptions for theory going forward.
On the flip side, future research may find a way to lift restrictions
that \SP{} currently adopts.

@@

\SP{} is a type checker and bytecode compiler.
It validates programs written in ordinary Python syntax
and generates type-specific bytecode for the Cinder runtime system.
Cinder is an extension of CPython 3.8 that adds a method-based
JIT compiler, virtual method tables, and several bytecode instructions.
In sum, the projects consists of both a new type system (\SP{})
and a new language (Cinder = Python extended to leverage types).

Running \SP{} code is no different than standard Python,
though it takes place on a different platform.
The main Cinder executable generates bytecode and performs
optimizations on the fly, same as CPython.
This is one of the main advantages of \SP{} over mypyc,
which implements type-directed optimizations for CPython
by compiling source code to C extension modules.

Almost all Python programs can run under \SP{} with no changes.
The few exceptions are due to restrictions that \SP{} puts
on classes to promote fast-running bytecode.
In general, switching to \SP{} requires minimal effort.
Any additional work that developers are willing to invest
in refining their types is rewarded by performance gains.

@@

\Cref{fig:sp-example} presents a first example program
that defines a function \code{f} and calls it from typed code.
The syntax is normal Python with type annotations.
Only the import statement makes special use of \SP{};
it imports a type \code{PyDict} that describes Python dictionaries.

Although this program is typed, a different untyped module is free
to import the function \code{f} and invoke it with any sort of
argument.
For this reason, the compiled version of \cref{fig:sp-example} must
include a few run-time casts:
\begin{itemize}
  \item
    The function \code{f} must check that its arguments match the \code{PyDict} domain annotation.
    Concretely, this check accepts any Python dictionary.
  \item
    Because the domain check does not validate the contents of an incoming dictionary,
    the dict access must check that its result matches the function codomain.
\end{itemize}
These checks ensure a kind of type soundness.

\SP{} also uses types to drive optimizations.
Because the call to \code{f} in \cref{fig:sp-example} appears in typed code,
there is no need to check that it sends type-correct inputs.
Thus, the compiled call uses a Cinder bytecode instruction to skip past the
domain check and immediately enter the function body.

The tension between checks that enforce soundness and optimizations allowed by soundness
is a central theme of \SP{}.

\subsection{Type Dynamic}

@@

\subsection{Checked Types}

@@

\subsection{Runtime Changes}

\begin{itemize}
  \item
    \textbf{VTables}
    Cinder adds virtual method tables to all classes.

\subsection{Code Restrictions}

\SP{} imposes several restrictions on classes and objects
to support efficient compilation:

\begin{itemize}
  \item \textbf{First-order Classes}.
    In order to benefit from types, class definitions must be defined at the top level of a module.
    Nested classes are supported by the language, but have type \code{Dynamic} according
    to the type checker.

  \item \textbf{No overloading}.
    Methods cannot be overloaded; each combination of a class
    and a name can refer to at most one method.
    This ensures that method resolution depends only on the
    type of the receiver object.
    % resolution does NOT depend on result type (T doesn't matter in `x:T = o.m()`)
    % Q. does resolution depend on args?

@@

\subsection{Type Restrictions}

Certain types are not supported:

@@

@@

@@

ctypes for primitives, unboxed representation

@@

\section{Related Work}

@@

>>>>

\begin{abstract}
  TBD
%  Soundness, migratory typing.
%  Novel run-time check strategy, combines concrete and transient for ergonomics.
\end{abstract}

@@

We have written this paper with two audiences in mind.
First, we want to encourage system-builders to reproduce the
\SP{} language design in another context and measure performance.
Second, we want to lend focus to theoretical research on gradual typing.
The restrictions that \SP{} adopts
may be worthwhile assumptions for theory going forward.
On the flip side, future research may find a way to lift some of
the restrictions currently in \SP{}.

@@

\SP{} is a type checker and bytecode compiler.
It validates programs written in ordinary Python syntax
and generates type-specific instructions for the Cinder runtime system.
The type system can express common Python idioms and is guaranteed
sound through run-time checks, which in turn enable optimizations.
The type system also includes a dynamic type as an escape hatch to
untyped behavior.
An unannotated variable behaves exactly as it would in Python.

Cinder is an extension of CPython 3.8 that adds a method-based
JIT compiler, virtual method tables, and several bytecode instructions
to directly express type checks and type-based optimizations.
These enhancements take full advantage of \SP{} type information;
they are built on top of CPython so that untyped code runs on
the platform.

Using \SP{} and Cinder is as easy as installing a new version
of CPython.
The main executable compiles programs to bytecode on the fly,
same as CPython, allowing standard developer tools to work.
By contrast, the only effective way to get type-directed optimizations
without changing CPython is to use C extension modules,
which introduce an extra compilation step and get in the way
of tooling~(\cref{s:related}).

@@

\Cref{fig:sp-example} presents a first example program.
It defines a function \code{f} and calls it from typed code.
The syntax is normal Python with type annotations.
Only the import statement makes special use of \SP{};
it imports a type \code{PyDict} that describes Python dictionaries.

Although this program is typed, a different untyped module is free
to import the function \code{f} and invoke it with any sort of
argument.
For this reason, the compiled version of \cref{fig:sp-example} must
include a few run-time casts:
\begin{itemize}
  \item
    The function \code{f} must check that its arguments match the \code{PyDict} domain annotation.
    Concretely, this check accepts any Python dictionary.
  \item
    Because the domain check does not validate the contents of an incoming dictionary,
    the dict access (\code{x["A"]}) must check that its result matches the function codomain.
\end{itemize}
These checks ensure a \emph{shape-level} notion of type soundness~(\cite{s:model}).
In short: if an expression has type \code{T} and reduces to a value, then the value
matches the top-level constructor, or shape, of the type.
The shape for a Python dictionary is merely \code{dict}.
Other types have deeper shapes; for example, \SP{} includes a \code{ChkDict}
type that predicts the keys and values in a special kind of dictionary.

Thanks to these checks, \SP{} can leverage type soundness to generate efficient code.
Because the call to \code{f} in \cref{fig:sp-example} appears in typed code,
there is no need to check that it sends type-correct inputs.
Thus, the compiled call uses a Cinder bytecode instruction to skip past the
domain check and immediately enter the function body.

The checks and optimizations in this example illustrate the general balancing act
of \SP{}.
Type soundness requires run-time checks, but also enables optimizations.
The gamble is that the optizations help more that the checks hurt.

\subsection{Type System Highlights}

\subsubsection{Type Dynamic}

@@

\subsubsection{Checked Types}

@@

\subsubsection{CTypes}

[FILL]
CTypes for primitives, unboxed representation.
Casts and unboxes happen at the boundary to Python.
Straightforward, not included in model.
Important for optimizations.

% https://github.com/facebookincubator/cinder/issues/52

\subsubsection{Restrictions}

OO restrictions:

\begin{itemize}
  \item \textbf{First-order Typed Classes}.
    In order to benefit from types, class definitions must be defined at the top level of a module.
    Nested classes are supported by the language, but have type \code{Dynamic} according
    to the type checker.

  \item \textbf{No Overloading}.
    Methods cannot be overloaded; each combination of a class
    and a name can refer to at most one method.
    This ensures that method resolution depends only on the
    type of the receiver object.
    % resolution does NOT depend on result type (T doesn't matter in `x:T = o.m()`)
    % Q. does resolution depend on args?

@@

Type restrictions:

@@

\subsection{Runtime System Highlights}

\begin{itemize}
  \item
    \textbf{VTables}
    Cinder adds virtual method tables to typed classes.
    ``The v-tables are only used for the \code{INVOKE\_METHOD} opcode, which
    the static compiler will only emit against a statically known static type.'' -Carl

%    No, there’s no overload resolution by arguments in Python. This difference
%    in behavior isn’t so much desired as just a consequence of making
%    `INVOKE_METHOD` optimizable. With a normal dynamic `CALL_FUNCTION`, first
%    the callable is placed on the stack, then the arguments, then there is a
%    `CALL_FUNCTION` (with number of args in oparg) to perform the call. This
%    means that first the callable is resolved, then the arguments. But with
%    `INVOKE_METHOD` we want to resolve the callable as part of the invoke
%    itself, since this gives us opportunity to inline-cache the target of the
%    call instead of always having to call something dynamic and unknown that’s
%    on the stack. So that necessarily implies that first the arguments are
%    resolved and placed on the stack, then the callable is resolved as part of
%    the `INVOKE_METHOD`. -Carl

@@

% things we did not model ... is that worth making a distinction here?

@@

@@

\section{Related Work}
\label{s:related}

@@

This is one of the main advantages of \SP{} over mypyc,
which implements type-directed optimizations for CPython
by compiling source code to C extension modules.
